---
title: Simulating differential HVGs 
author: Aaron Lun
date: 3 February 2017
output:
  html_document:
    fig_caption: false 
---

```{r, echo=FALSE, results="hide"}
dir.create("figure-diff", showWarning=FALSE)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE, fig.path="figure-diff/")
options(width=100)
```

# Overview

The problem is that changes to technical noise do not preserve the measures of biological variability.
We will consider the following model:

- True transcript molecule counts are distributed based on a NB distribution, with mean $\mu$ and _biological_ dispersion $\phi$.
- Transcripts are captured into cDNA molecules based on a binomial distribution with probability $p$, itself Beta-distributed across cells.
- The number of cDNA molecules is scaled by $(1+X)^3$ where $X$ is Beta-distributed across cells and represents amplification efficiency across 3 PCR cycles.
- Seqeuncing noise is binomial with some constant probability across cells (we will ignore fragmentation for now).

This yields the following simulation function:

```{r}
ncells <- 10000
set.seed(105872)
simCounts <- function(mu, disp, capture.args, amp.args, seq.prop) {
    y.t <- rnbinom(ncells, mu=mu, size=1/disp)
    y.c <- rbinom(ncells, size=y.t, p=rbeta(ncells, capture.args[1], capture.args[2]))
    amp.eff <- (1+rbeta(ncells, amp.args[1], amp.args[2]))^3
    y.ca <- round(y.c * amp.eff)
    y.r <- rbinom(ncells, size=y.ca, p=seq.prop)
    return(y.r)
}
```

We also define various metrics for computing the biological component of variation.

```{r}
getBio <- function(bio, tech) {
    bio.cv2 <- var(bio)/mean(bio)^2
    tech.cv2 <- var(tech)/mean(tech)^2
    bio.log <- var(log2(bio+1)) 
    tech.log <- var(log2(tech+1))
    return(list(Estimates=data.frame(bio.cv2, tech.cv2, bio.log, tech.log),
                Components=data.frame(AbsCV2=bio.cv2 - tech.cv2,
                                      RelCV2=bio.cv2/tech.cv2,
                                      AbsLog=bio.log - tech.log,
                                      RelLog=bio.log/tech.log)))
}
```

To demonstrate, let's set up a simulation involving zero and non-zero values of `disp`.

```{r}
mu <- c(10, 10, 100, 100, 1000, 1000)
disp <- c(0.5, 2, 0.5, 2, 0.5, 2)
cap.args <- c(5, 50)
amp.args <- c(10, 1)
seq.prop <- 0.1
```

Now we compute biological components across all simulation scenarios.

```{r simulate_and_run}
collected.com <- list()
collected.est <- list()
for (i in seq_along(mu)) {
    y.bio <- simCounts(mu=mu[i], disp=disp[i], capture.args=cap.args, amp.args=amp.args, seq.prop=seq.prop)
    y.tech <- simCounts(mu=mu[i], disp=1e-8, capture.args=cap.args, amp.args=amp.args, seq.prop=seq.prop)
    out <- getBio(y.bio, y.tech)
    collected.est[[i]] <- data.frame(Mu=mu[i], Disp=disp[i], out$Estimates)
    collected.com[[i]] <- data.frame(Mu=mu[i], Disp=disp[i], out$Components)
}
do.call(rbind, collected.est)
obs <- do.call(rbind, collected.com)
obs
```

Storing these as the reference results.

```{r}
ref <- obs
``` 

# Changing the capture efficiency

First, we consider what happens if capture efficiency increases.

```{r}
cap.args <- c(10, 50)
amp.args <- c(10, 1)
```

Running across the simulations.

```{r, ref.label="simulate_and_run"}
```

Calculating the relative difference (in percentages) from the reference.

```{r}
(obs/ref - 1)*100
```

# Changing the amplification rate

Increasing the variability of the amplification distribution.

```{r}
cap.args <- c(5, 50)
amp.args <- c(5, 1)
```

Running across the simulations.

```{r, ref.label="simulate_and_run"}
```

Calculating the relative difference from the reference.

```{r}
(obs/ref - 1)*100
```

# Changing the sequencing depth

Increasing the sequencing depth.

```{r}
cap.args <- c(5, 50)
amp.args <- c(10, 1)
seq.prop <- 0.5
```

Running across the simulations.

```{r, ref.label="simulate_and_run"}
```

Calculating the relative difference from the reference.

```{r}
(obs/ref - 1)*100
```

# Conclusion

Of the possible metrics for computing biological components, the absolute difference in the CV^2^ is most stable with respect to changes in the technical noise profile.
Note that this only accounts for the expectation of the biological component and does not consider changes to the sampling distribution.
It also makes no statement about the validity of the null model, either, for any given scenario.
We're focusing here on distortions to the size of the biological component as this will affect detection differential variability.

